Note on a noisy variant of the RMD

### 2016-10-16 mf_stuff/note-on-noisy-RMD/note-on-noisy-RMD.tex
extension model sketch final; minor tweaks to extension model code

Our present model computes the mutation probability that a teacher type $t_i$ has
offspring of type $t_j$ as:

\begin{align*}
  Q_{ij}  \propto \sum_{d \in D} P(d|t_i) F(t_j,d) \,, \ \  \text{where} \ \
F(t_j,d)  \propto P(t_j|d)^l \ \  \text{and} \ \ P(t_j|d) \propto P(t_j) P(d|t_j)
\end{align*}

@@

>>>>

Our present model computes the mutation probability that a teacher type $t_i$ has
offspring of type $t_j$ as:
\begin{align*}
  Q_{ij}  \propto \sum_{d \in D} P(d|t_i) F(t_j,d) \,, \ \  \text{where} \ \
F(t_j,d)  \propto P(t_j|d)^l \ \  \text{and} \ \ P(t_j|d) \propto P(t_j) P(d|t_j)
\end{align*}

@@

\subsection*{RMD with noisy perception}

This model is implemented in file \texttt{3rmd-singlescalar\_withNoise.py} in the new repository
\texttt{rmd-bounds2}.

We have two states: $s_\exists$ (some but not all) and $s_\forall$ (all). Let's assume that the
probabilities for observing the state in the column when the actual state is the state in the
row are given by this table:
\begin{center}
  \begin{tabular}{ccc}
    & $s_\exists$ & $s_\forall$ \\ \midrule 
    $s_\exists$ & $1-\epsilon$ & $\epsilon$ \\
    $s_\forall$ & $\delta$ & $1-\delta$
  \end{tabular}
\end{center}
So, $\epsilon$ is the error probability when the true state is $s_\exists$ and $\delta$ is the
error probability when the true state is $s_\forall$. How does this affect our model of
Bayesian learning?

Let's denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the
actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$ ). This is given by the table
above. We can then derive the following probabilities. Firstly, the probability that $s_a$ is
the actual state when the learner observes $s_l$ is just Bayes rule, combining prior of $s_a$
with the noise from above:
\begin{align*}
  P_N(s_a \mid s_l) \propto P(s_a) \ P_N(s_l \mid s_a)\,.
\end{align*}
Secondly, the probability that the teacher observes $s_t$ when the learner observes $s_l$ is:
\begin{align*}
  P_N(s_t \mid s_l) = \sum_{s_a} P(s_a \mid s_l) \ P_N(s_t \mid s_a)\,.
\end{align*}
Finally, this gives us the probability that a teacher of type $t$ produces a datum that is
perceived by the listener as $d = \tuple{s_l, m}$:
\begin{align*}
  P_N(\tuple{s_l, m} \mid t) = \sum_{s_t} P_N(s_t \mid s_l) \ P(m \mid s_t, t)\,.
\end{align*}
Generalize this to a sequence of perceived data $d_l$ and write $P_N(d_l \mid t)$. We then
define the noise-perturbed mutation matrix as:
\begin{align*}
  Q_{ij}  \propto \sum_{d_l \in D} P(d_l \mid t_i) F(t_j,d_l) \,, \ \  \text{where $F(t_j,d)$
    as before.}
\end{align*}
In words, it may be the case that learner and/or teacher do not perceive the actual state as
what it is. They are not aware of this, and produce/learn as if what they observed was the
actual state. In particular, the learner does not reason about noise when he tries to infer the
speaker's type. He takes what he observes a state to be as the actual state that the teacher
has seen as well and computes which type would have most likely generated the message to this
state. This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some
types err in a way that resembles the noiseless behavior of other types. I.e., this could, in
principle, induce transmission biases that look as is there was a cognitive bias in favor of a
particular type, simply because that type better explains the noise.

On top of changing the mutation matrix in this way, we also need to adapt the calculation of
expected utilities, taking into consideration that states are perceived noisily. So, where
before we had:
\begin{align*}
  U_S(t_i,t_j) = \sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,,
\end{align*}
we now have:
\begin{align*}
  U_S(t_i,t_j) = \sum_{s_a}  P(s_a) \sum_{s_t} P_N(s_t \mid s_a) \sum_m S_n(m|s_t;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,.
\end{align*}

\subsection*{Results}

In a nutshel: this works for some parameter settings but not for others. I have not fully
explored the whole space. I would be content with a paper that says: there are parameter values
that produce behavior as if there are cognitive biases without cognitive biases. I believe that
this would be an additional and very interesting contribution, even if there are noise
structures that have other effects.

One parameter setting that gives us the desired result (most prevalent type is $t_{11}$) is:
\begin{verbatim}
alpha = 5 # rate to control difference between semantic and pragmatic violations
cost = 0 # cost for LOT-concept with upper bound
lam = 20 # soft-max parameter
k = 3  # length of observation sequences
sample_amount = 50 #amount of k-length samples for each production type
epsilon = 0.3 # probability of perceiving S-all, when true state is S-sbna
delta = 0.1 # probability of perceiving S-sbna, when true state is S-all
gens = 20 #number of generations per simulation run
runs = 50 #number of independent simulation runs
states = 2 #number of states
messages = 2 #number of messages
learning_parameter = 10 #prob-matching = 1, increments approach MAP
\end{verbatim}
One sample result is these parameter values is:
\begin{verbatim}
[ 0.01 ,  0.01 ,  0.01 ,  0.012,  0.028,  0.024,  0.01 ,  0.01 ,
        0.01 ,  0.301,  0.471,  0.102]
\end{verbatim}
Importantly, we don't need costs ($c = 0$)! Unfortunately, we do
need to tinker with $\alpha$ (for defensible, techical reasons) and we need to assume that
$\epsilon > \delta$, otherwise either $t_{10}$ or even $t_{12}$ may dominate.

That means that if we wanted to also argue that this is a plausible explanation for
lower-bounds semantics, we'd need to argue why it is natural that $\epsilon > \delta$. Maybe
such a story can be given, but I presently do not see why/how any such story is obviously
better than any other story I could come up with for any other relation of $\epsilon$ and
$\delta$.

One reasoning for $\epsilon > \delta$ could be this. When the true state is $s_a$, we would
actually need negative evidence that some item/person/whatever does not have a
property. Negative evidence of this kind might be relatively infrequent, especially if
untrue. On the other hand, perceiving $s_a$ when the true state is $s_e$ could be more
frequent, because it would, for example, require wrong information about the domain size. It
could also be a cause of overemphasizing: speakers tend to exaggerate and want to claim
\emph{all}, when in fact that is not true \citep[e.g.][for a model that has the speakers'
tendency to overemphasize as a motor of language change]{Schaden2012:Modelling-the-A}.
